---
title: "The Shoe Price Gouging Hypothesis"
subtitle: "How a tracking cookie might be picking your wallet"
author: Ben Graul
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    theme: cosmo
    fig-width: 8
    fig-height: 5
execute:
  warning: false
  message: false
---

## The Setup

You're shopping for shoes online. The site has decided you're female. You start
to wonder whether your friend on his own laptop sees the same prices. That is
the **Shoe Price Gouging Hypothesis** in plain English: the store reads your
tracking cookie and quietly nudges prices around based on who it thinks you are.

We're going to investigate using exactly what we covered in class:
independent-samples *t*-tests, one-sample *z*-tests, paired *t*-tests, and
**one-way ANOVA**. Plots throughout, so you can see what the numbers are doing.

A heads-up about the figures: every plot shows 95% confidence intervals on the
means. Bring questions about what those intervals actually mean to office hours. 
The short version for now: wider interval, less certainty about where
the true mean sits.

## The Data

We collected prices under two conditions. First with the cookie identifying us
as female, then again after clearing cookies and letting the site assume male.

```{r setup}
library(tidyverse)
library(ggsignif)

# Population parameters from the full inventory
mu_men      <- 86.74
sigma_men   <- 13.86
mu_women    <- 95.70
sigma_women <- 16.13

shoe_data <- tribble(
  ~shoe,     ~cookie, ~size,    ~mean,  ~sd,    ~n,
  "Men's",   "F",     "Small",  103.80, 10.59,  5,
  "Men's",   "F",     "Large",  88.14,  14.75,  100,
  "Men's",   "M",     "Small",  79.37,  8.28,   5,
  "Men's",   "M",     "Large",  84.73,  14.13,  100,
  "Women's", "F",     "Small",  102.85, 29.80,  5,
  "Women's", "F",     "Large",  95.93,  16.52,  100,
  "Women's", "M",     "Small",  62.32,  1.58,   5,
  "Women's", "M",     "Large",  95.52,  15.80,  100
) |>
  mutate(
    se    = sd / sqrt(n),
    ci_lo = mean - qt(0.975, n - 1) * se,
    ci_hi = mean + qt(0.975, n - 1) * se
  )

shoe_data
```

Quick read on the populations: men's shoes really do average about \$87, women's
shoes really do average about \$96, and that gap of roughly \$9 is the truth
we're trying to recover from samples.

## Round 1: Independent-Samples *t*-Tests (Cookie = F)

### Test 1: Five Shoes Each

We grabbed five men's shoes and five women's shoes. Equal variances assumed,
since that's what we covered.

```{r small-t}
two_sample_t <- function(M1, SD1, n1, M2, SD2, n2) {
  pooled_var <- ((n1 - 1) * SD1^2 + (n2 - 1) * SD2^2) / (n1 + n2 - 2)
  pooled_sd  <- sqrt(pooled_var)
  se         <- pooled_sd * sqrt(1/n1 + 1/n2)
  t_stat     <- (M1 - M2) / se
  df         <- n1 + n2 - 2
  p_val      <- 2 * (1 - pt(abs(t_stat), df))
  tibble(t = t_stat, df = df, p = p_val,
         pooled_sd = pooled_sd, se = se)
}

test_small <- two_sample_t(103.80, 10.59, 5, 102.85, 29.80, 5)
test_small
```

`t(8) = 0.07, p = .948.`

That \$29.80 standard deviation on the women's side is doing the work here.
With five shoes and that much spread, the test has nothing to grab onto.

### Test 2: 100 Shoes Each

Same store, same cookie, just more shoes.

```{r large-t}
test_large <- two_sample_t(88.14, 14.75, 100, 95.93, 16.52, 100)
test_large
```

`t(198) = -3.52, p < .001.`

There it is. Women's shoes really do cost about \$8 more on average. The effect
was always there. Five shoes just couldn't see it.

### Visualizing the Power Problem

```{r power-plot, fig.height=5.5}
plot_round1 <- shoe_data |>
  filter(cookie == "F") |>
  mutate(size = factor(size,
                       levels = c("Small", "Large"),
                       labels = c("n = 5 each", "n = 100 each")))

# Bracket positions for significance annotations
brackets_round1 <- tibble(
  size      = factor(c("n = 5 each", "n = 100 each"),
                     levels = c("n = 5 each", "n = 100 each")),
  y_pos     = c(160, 125),
  label     = c("p = .948 (ns)", "p < .001 ***")
)

ggplot(plot_round1, aes(x = shoe, y = mean, fill = shoe)) +
  geom_col(width = 0.6, alpha = 0.85) +
  geom_errorbar(aes(ymin = ci_lo, ymax = ci_hi), width = 0.2, linewidth = 0.7) +
  geom_signif(
    data = data.frame(size = factor("n = 5 each",
                                    levels = c("n = 5 each", "n = 100 each"))),
    aes(xmin = "Men's", xmax = "Women's",
        annotations = "p = .948 (ns)", y_position = 155),
    inherit.aes = FALSE, manual = TRUE, tip_length = 0.02
  ) +
  geom_signif(
    data = data.frame(size = factor("n = 100 each",
                                    levels = c("n = 5 each", "n = 100 each"))),
    aes(xmin = "Men's", xmax = "Women's",
        annotations = "p < .001 ***", y_position = 110),
    inherit.aes = FALSE, manual = TRUE, tip_length = 0.02
  ) +
  facet_wrap(~ size) +
  scale_fill_manual(values = c("Men's" = "#2E86AB", "Women's" = "#C8467F")) +
  scale_y_continuous(labels = scales::dollar_format(), limits = c(0, 175)) +
  labs(
    x = NULL, y = "Mean price",
    title = "Same store, same cookie, same gender gap",
    subtitle = "Five shoes can't see what 100 shoes can. Bars show 95% CIs."
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold"))
```

The bars are similar heights in both panels. What changes is the width of the
confidence intervals. Five shoes give you huge intervals that overlap massively.
A hundred shoes give you tight intervals that clearly separate. The data didn't
change. Our ability to see the data did.

## Round 2: One-Sample *z*-Tests

We have the population standard deviations, so a *z*-test is appropriate. We're
going to ask three questions about the curated samples we got served.

```{r z-helper}
one_sample_z <- function(M, mu, sigma, n) {
  se <- sigma / sqrt(n)
  z  <- (M - mu) / se
  p  <- 2 * (1 - pnorm(abs(z)))
  tibble(z = z, p = p, se = se)
}
```

### Test 3: Is the Curated Men's Sample Weird Compared to All Men's Shoes?

```{r z1}
z_men_vs_men <- one_sample_z(M = 103.80, mu = mu_men,
                              sigma = sigma_men, n = 5)
z_men_vs_men
```

`z = 2.75, p = .006.` So the five "men's shoes" the algorithm picked for the
female-cookie shopper aren't a normal slice of the men's inventory. They're
significantly more expensive than the average men's shoe.

### Test 4: Does That Same Curated Sample Look Like Women's Shoes?

```{r z2}
z_men_vs_women <- one_sample_z(M = 103.80, mu = mu_women,
                                sigma = sigma_women, n = 5)
z_men_vs_women
```

`z = 1.12, p = .261.` The "men's" shoes shown to a female-cookie shopper are
priced indistinguishably from the women's population. The algorithm isn't
picking random men's shoes. It's picking men's shoes that cost what women's
shoes cost.

### Test 5: Sanity Check on the Large Women's Sample

```{r z3}
z_women_check <- one_sample_z(M = 95.93, mu = mu_women,
                               sigma = sigma_women, n = 100)
z_women_check
```

`z = 0.14, p = .886.` Sanity check passed. When we sample broadly, we recover
the population. The bias lives in the small, curated picks.

### Visualizing Where the Curated Sample Falls

```{r z-plot, fig.height=5}
x_grid <- seq(40, 150, length.out = 400)
densities <- bind_rows(
  tibble(price = x_grid,
         density = dnorm(x_grid, mu_men, sigma_men),
         population = "Men's population"),
  tibble(price = x_grid,
         density = dnorm(x_grid, mu_women, sigma_women),
         population = "Women's population")
)

# Distribution of the SAMPLE MEAN under each null hypothesis
sample_mean_curves <- bind_rows(
  tibble(price = x_grid,
         density = dnorm(x_grid, mu_men, sigma_men/sqrt(5)),
         hypothesis = "If men's pop"),
  tibble(price = x_grid,
         density = dnorm(x_grid, mu_women, sigma_women/sqrt(5)),
         hypothesis = "If women's pop")
)

ggplot(sample_mean_curves, aes(x = price, y = density, color = hypothesis)) +
  geom_line(linewidth = 1.1) +
  geom_vline(xintercept = 103.80, linetype = "dashed",
             color = "#222222", linewidth = 0.7) +
  annotate("label", x = 103.80, y = 0.058,
           label = "Curated men's\nsample mean = $103.80",
           hjust = 0.5, size = 3.3, fill = "white") +
  annotate("text", x = 78, y = 0.045,
           label = "z = 2.75\np = .006",
           color = "#2E86AB", size = 3.3, fontface = "bold") +
  annotate("text", x = 118, y = 0.025,
           label = "z = 1.12\np = .261",
           color = "#C8467F", size = 3.3, fontface = "bold") +
  scale_color_manual(values = c("If men's pop" = "#2E86AB",
                                "If women's pop" = "#C8467F"),
                     name = "Sampling distribution") +
  scale_x_continuous(labels = scales::dollar_format()) +
  labs(
    x = "Sample mean ($, n = 5)", y = "Density",
    title = "The curated sample sits far from men's, comfortable with women's",
    subtitle = "Sampling distributions of the mean for n = 5"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold"))
```

The dashed line marks the curated sample mean of \$103.80. Look how far it sits
in the right tail of the men's sampling distribution, and how comfortably it
sits inside the women's sampling distribution. Visually, the *z*-tests are just
reading off where that line falls.

## Round 3: The Plot Twist (Cookie = M)

We cleared cookies, browsed a few hardware sites, and came back. Now the algorithm
thinks we're male. We refreshed the same shoes.

```{r cookie-comparison}
shoe_data |>
  select(shoe, cookie, size, mean) |>
  pivot_wider(names_from = cookie, values_from = mean,
              names_prefix = "Cookie = ") |>
  mutate(`Difference (F - M)` = `Cookie = F` - `Cookie = M`)
```

Female-cookie prices are higher across the board, and the gap is huge for the
small curated samples. The right test for this is paired, since each cookie=F
price has a matched cookie=M price for the same physical shoe.

## Round 4: Paired *t*-Test

For the paired test, we use the differences directly. Across the 100 men's
shoes, the average price drop after switching to a male cookie was \$3.41,
with a standard deviation of differences of \$9.20. For the women's shoes, the
average drop was \$0.41 with SD of differences \$4.51.

```{r paired-t}
paired_t <- function(mean_diff, sd_diff, n) {
  se <- sd_diff / sqrt(n)
  t_stat <- mean_diff / se
  df <- n - 1
  p <- 2 * (1 - pt(abs(t_stat), df))
  tibble(mean_diff = mean_diff, sd_diff = sd_diff, n = n,
         t = t_stat, df = df, p = p)
}

paired_results <- bind_rows(
  paired_t(mean_diff = 3.41, sd_diff = 9.20, n = 100) |>
    mutate(comparison = "Men's shoes (F vs M cookie)"),
  paired_t(mean_diff = 0.41, sd_diff = 4.51, n = 100) |>
    mutate(comparison = "Women's shoes (F vs M cookie)")
) |>
  select(comparison, everything())

paired_results
```

For the men's shoes, *t*(99) = 3.71, *p* < .001. For the women's shoes,
*t*(99) = 0.91, *p* = .365. The cookie effect is real for men's shoes and
absent for women's shoes. The store appears to be charging female-coded
shoppers more specifically when they shop in the men's section.

```{r paired-plot, fig.height=4.5}
paired_plot_data <- shoe_data |>
  filter(size == "Large") |>
  select(shoe, cookie, mean, ci_lo, ci_hi)

ggplot(paired_plot_data, aes(x = cookie, y = mean, group = shoe, color = shoe)) +
  geom_line(linewidth = 1.2) +
  geom_errorbar(aes(ymin = ci_lo, ymax = ci_hi), width = 0.08, linewidth = 0.7) +
  geom_point(size = 4) +
  geom_text(aes(label = scales::dollar(mean, accuracy = 0.01)),
            nudge_y = 2, size = 3.5, show.legend = FALSE) +
  annotate("text", x = 1.5, y = 90,
           label = "Men's: t(99) = 3.71, p < .001 ***",
           size = 3.5, color = "#2E86AB", fontface = "bold") +
  annotate("text", x = 1.5, y = 99,
           label = "Women's: t(99) = 0.91, p = .365 (ns)",
           size = 3.5, color = "#C8467F", fontface = "bold") +
  scale_color_manual(values = c("Men's" = "#2E86AB", "Women's" = "#C8467F")) +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(
    x = "Tracking cookie", y = "Mean price",
    title = "The cookie effect is selective",
    subtitle = "Men's prices drop when the cookie says male. Women's prices barely move."
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold"))
```

The men's line has a visible slope. The women's line is nearly flat. The
tracking cookie is changing prices in a targeted way.

## Round 5: One-Way ANOVA

We have four groups: each combination of shoe gender and cookie status. A
one-way ANOVA treats them as four conditions and asks whether at least one mean
differs from the others.

```{r anova}
anova_groups <- shoe_data |>
  filter(size == "Large") |>
  mutate(group = paste0(shoe, ", Cookie=", cookie)) |>
  select(group, mean, sd, n)

anova_groups

# Compute one-way ANOVA from group summary statistics
k          <- nrow(anova_groups)
N          <- sum(anova_groups$n)
grand_mean <- with(anova_groups, sum(mean * n) / N)

SS_between <- with(anova_groups, sum(n * (mean - grand_mean)^2))
SS_within  <- with(anova_groups, sum((n - 1) * sd^2))

df_between <- k - 1
df_within  <- N - k
MS_between <- SS_between / df_between
MS_within  <- SS_within  / df_within
F_stat     <- MS_between / MS_within
p_value    <- 1 - pf(F_stat, df_between, df_within)

anova_table <- tribble(
  ~Source,           ~SS,                       ~df,        ~MS,                       ~F,                  ~p,
  "Between groups",  round(SS_between, 2),      df_between, round(MS_between, 2),      round(F_stat, 2),    formatC(p_value, format = "e", digits = 2),
  "Within groups",   round(SS_within, 2),       df_within,  round(MS_within, 2),       NA_real_,            "",
  "Total",           round(SS_between + SS_within, 2), df_between + df_within, NA_real_, NA_real_, ""
)
anova_table
```

`F(3, 396) = 13.08, p < .001.` At least one of the four conditions has a
different mean than the others. The ANOVA on its own won't tell us which pairs
differ. We can see the pattern in the means though.

```{r anova-plot, fig.height=5.5}
anova_plot_data <- shoe_data |>
  filter(size == "Large") |>
  mutate(group = paste0(shoe, ",\nCookie=", cookie),
         group = factor(group, levels = c("Men's,\nCookie=F",
                                          "Men's,\nCookie=M",
                                          "Women's,\nCookie=F",
                                          "Women's,\nCookie=M")))

ggplot(anova_plot_data, aes(x = group, y = mean, fill = shoe)) +
  geom_col(width = 0.65, alpha = 0.85) +
  geom_errorbar(aes(ymin = ci_lo, ymax = ci_hi),
                width = 0.18, linewidth = 0.7) +
  geom_text(aes(label = scales::dollar(mean, accuracy = 0.01)),
            vjust = -0.6, size = 3.4, fontface = "bold") +
  geom_signif(
    annotations = "Men's: F vs M cookie, p < .001 ***",
    xmin = "Men's,\nCookie=F", xmax = "Men's,\nCookie=M",
    y_position = 110, tip_length = 0.02, textsize = 3.2
  ) +
  geom_signif(
    annotations = "Women's: F vs M cookie, p = .365 (ns)",
    xmin = "Women's,\nCookie=F", xmax = "Women's,\nCookie=M",
    y_position = 118, tip_length = 0.02, textsize = 3.2
  ) +
  annotate("label", x = 2.5, y = 130,
           label = "Overall ANOVA: F(3, 396) = 13.08, p < .001",
           fill = "#FFF3B0", size = 3.6, fontface = "bold") +
  scale_fill_manual(values = c("Men's" = "#2E86AB", "Women's" = "#C8467F")) +
  scale_y_continuous(labels = scales::dollar_format(),
                     limits = c(0, 140)) +
  labs(
    x = NULL, y = "Mean price",
    title = "Four groups, one ANOVA, a clear pattern",
    subtitle = "95% CIs shown. Brackets are paired comparisons from Round 4."
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold"),
        axis.text.x = element_text(size = 9))
```

Two stories show up in the bars. Women's shoes are taller than men's shoes
regardless of cookie. Within men's shoes, the female-cookie bar is taller than
the male-cookie bar. Within women's shoes, the cookie barely matters. That's
the pattern the ANOVA detected.

## So, Is the Store Gouging?

Pulling everything together:

- The **large-sample independent *t*-test** showed women's shoes really do cost
  more than men's shoes, by about \$8.
- The ***z*-tests** showed the curated small sample of men's shoes shown to a
  female-cookie shopper was priced up in women's-shoe territory. The algorithm
  was cherry-picking expensive men's shoes for that audience.
- The **paired *t*-test** showed that on the same men's shoes, female-cookie
  shoppers pay about \$3.41 more than male-cookie shoppers. For women's shoes,
  the cookie does nothing.
- The **one-way ANOVA** confirmed the four shoe-by-cookie conditions aren't
  drawn from the same distribution.

The hypothesis isn't proven beyond doubt. A real audit would need a controlled
experiment with many more shoes, randomized browsing, and a few different IP
addresses. The evidence we have is consistent and points one direction though.
The cookie matters, and sample size matters even more. With five shoes, none of
this would have been visible.

::: {.callout-tip}
## The Big Lesson
A small biased sample can perfectly hide a large biased system. Always ask how
much data a conclusion came from.
:::

::: {.callout-note}
## For the Review Session
Bring questions about confidence intervals. Every plot in this document shows
them, and we'll spend time talking about what a 95% CI does and does not say.
:::
