---
title: "The Shoe Price Gouging Hypothesis"
subtitle: "How a tracking cookie might be picking your wallet"
author: "Ben Graul"
date: today
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    code-fold: show
    theme: flatly
    fig-width: 8
    fig-height: 5
execute:
  warning: false
  message: false
---

## The Setup

You're shopping for shoes online. The site has decided you're female. You start
to wonder whether your friend on his own laptop sees the same prices. That is
the **Shoe Price Gouging Hypothesis** in plain English: the store reads your
tracking cookie and quietly nudges prices around based on who it thinks you are.

We're going to investigate using exactly what we covered in class:
independent-samples *t*-tests, one-sample *z*-tests, paired *t*-tests, and
**one-way ANOVA**. Plots throughout, so you can see what the numbers are doing.

A heads-up about the figures: every plot shows 95% confidence intervals on the
means. Bring questions about what those intervals actually mean to the review
session. The short version for now: wider interval, less certainty about where
the true mean sits.


```{r setup}
library(tidyverse)
library(kableExtra)
library(ggsignif)

theme_set(theme_minimal(base_size = 12) +
            theme(plot.title = element_text(face = "bold")))

shoe_palette <- c("Men's" = "#2E86AB", "Women's" = "#C8467F")

# Read population data from CSV with tidy column names
pop_raw <- read_csv("shoe_prices_sampling_numbers.csv",
                     col_names = c("mens_pop", "womens_pop", "mens_pop_m", "womens_pop_m"),
                     skip = 1)

# Extract and clean population vectors (remove $ and convert to numeric)
mens_pop <- pop_raw |>
  pull(mens_pop) |>
  str_remove_all("\\$|,") |>
  as.numeric() |>
  na.omit()

womens_pop <- pop_raw |>
  pull(womens_pop) |>
  str_remove_all("\\$|,") |>
  as.numeric() |>
  na.omit()

# Calculate population parameters
mu_men      <- mean(mens_pop)
sigma_men   <- sd(mens_pop)
mu_women    <- mean(womens_pop)
sigma_women <- sd(womens_pop)
```

### Simulating the Cookie Effect

The store's recommendation algorithm "curates" what you see. We model that as
selection bias: when the cookie says female, the algorithm pulls men's shoes
from the upper end of the inventory. When the cookie says male, it pulls from
the lower end. Large samples (you scrolled through 100 shoes) are too big to
fully curate, so they look mostly representative, but a small dynamic discount
is applied to male-cookie users.

```{r simulate}
set.seed(110)

mens_sorted_desc   <- sort(mens_pop,   decreasing = TRUE)
mens_sorted_asc    <- sort(mens_pop)
womens_sorted_asc  <- sort(womens_pop)

# Small samples: algorithm has time to curate
mens_small_F   <- sample(mens_sorted_desc[1:30], 5)   # top of men's inventory
mens_small_M   <- sample(mens_sorted_asc[1:30],  5)   # bottom of men's inventory
womens_small_F <- sample(womens_pop, 5)               # roughly random
womens_small_M <- sample(womens_sorted_asc[1:30], 5)  # bottom of women's inventory

# Large samples: representative, with a small discount applied for cookie = M
mens_large_F   <- sample(mens_pop, 100)
mens_large_M   <- sample(mens_pop, 100) - 3.41
womens_large_F <- sample(womens_pop, 100)
womens_large_M <- sample(womens_pop, 100) - 0.41

# Tidy summary table for plotting
shoe_data <- tribble(
  ~shoe,     ~cookie, ~size,    ~values,
  "Men's",   "F",     "Small",  mens_small_F,
  "Men's",   "F",     "Large",  mens_large_F,
  "Men's",   "M",     "Small",  mens_small_M,
  "Men's",   "M",     "Large",  mens_large_M,
  "Women's", "F",     "Small",  womens_small_F,
  "Women's", "F",     "Large",  womens_large_F,
  "Women's", "M",     "Small",  womens_small_M,
  "Women's", "M",     "Large",  womens_large_M
) |>
  mutate(
    mean  = map_dbl(values, mean),
    sd    = map_dbl(values, sd),
    n     = map_int(values, length),
    se    = sd / sqrt(n),
    ci_lo = mean - qt(0.975, n - 1) * se,
    ci_hi = mean + qt(0.975, n - 1) * se
  )

shoe_data |> select(-values)
kable(shoe_data)
```

## Round 1: Independent-Samples *t*-Tests (Cookie = F)


### Test 1: Five Shoes Each

We grabbed five men's shoes and five women's shoes. Equal variances assumed,
since that's what we covered.

```{r small-t}
t_small_F <- t.test(mens_small_F, womens_small_F, 
                    var.equal = TRUE, 
                    alternative = "two.sided")

# Extract components for reporting
t_stat_small <- t_small_F$statistic
df_small <- t_small_F$parameter
p_small <- t_small_F$p.value

# Calculate Cohen's d for effect size
pooled_sd_small <- sqrt(((length(mens_small_F) - 1) * var(mens_small_F) + 
                          (length(womens_small_F) - 1) * var(womens_small_F)) /
                         (length(mens_small_F) + length(womens_small_F) - 2))
cohens_d_small <- (mean(mens_small_F) - mean(womens_small_F)) / pooled_sd_small

tibble(
  t = t_stat_small,
  df = df_small,
  p = p_small,
  `Cohen's d` = cohens_d_small,
  `95% CI (d)` = sprintf("[%.2f, %.2f]", 
                         cohens_d_small - 1.96 * sqrt(2/length(mens_small_F)), 
                         cohens_d_small + 1.96 * sqrt(2/length(mens_small_F)))
)
```

Men's shoes came in at *M* = \$`r round(mean(mens_small_F), 2)`,
*SD* = \$`r round(sd(mens_small_F), 2)`. Women's shoes came in at
*M* = \$`r round(mean(womens_small_F), 2)`,
*SD* = \$`r round(sd(womens_small_F), 2)`. The test gave
*t*(`r df_small`) = `r round(t_stat_small, 2)`,
*p* = `r round(p_small, 3)`, *d* = `r round(cohens_d_small, 2)`.

Even if we had a difference (which we don't quite see), Cohen's *d* = `r round(cohens_d_small, 2)`
would indicate a `r if(abs(cohens_d_small) < 0.2) "negligible" else if(abs(cohens_d_small) < 0.5) "small" else if(abs(cohens_d_small) < 0.8) "medium" else "large"` 
effect size. That women's-side standard deviation is doing the work. With five shoes and
that much spread, the test has nothing to grab onto.

### Test 2: 100 Shoes Each

Same store, same cookie, more shoes.

```{r large-t}
t_large_F <- t.test(mens_large_F, womens_large_F, 
                    var.equal = TRUE, 
                    alternative = "two.sided")

# Extract components for reporting
t_stat_large <- t_large_F$statistic
df_large <- t_large_F$parameter
p_large <- t_large_F$p.value
ci_large <- t_large_F$conf.int

# Calculate Cohen's d for effect size
pooled_sd_large <- sqrt(((length(mens_large_F) - 1) * var(mens_large_F) + 
                          (length(womens_large_F) - 1) * var(womens_large_F)) /
                         (length(mens_large_F) + length(womens_large_F) - 2))
cohens_d_large <- (mean(mens_large_F) - mean(womens_large_F)) / pooled_sd_large

tibble(
  t = t_stat_large,
  df = df_large,
  p = p_large,
  `Cohen's d` = cohens_d_large,
  `95% CI (d)` = sprintf("[%.2f, %.2f]", 
                         cohens_d_large - 1.96 * sqrt(2/length(mens_large_F)), 
                         cohens_d_large + 1.96 * sqrt(2/length(mens_large_F)))
)
```

Men's came in at *M* = \$`r round(mean(mens_large_F), 2)`,
*SD* = \$`r round(sd(mens_large_F), 2)`. Women's at
*M* = \$`r round(mean(womens_large_F), 2)`,
*SD* = \$`r round(sd(womens_large_F), 2)`.
*t*(`r df_large`) = `r round(t_stat_large, 2)`,
*p* = `r format.pval(p_large, digits = 2, eps = 0.001)`,
*d* = `r round(cohens_d_large, 2)`.

There it is. Women's shoes really do cost more on average, with a 
**`r if(abs(cohens_d_large) < 0.2) "negligible" else if(abs(cohens_d_large) < 0.5) "small" else if(abs(cohens_d_large) < 0.8) "medium" else "large"` effect size**.
The effect was always there. Five shoes just couldn't reveal it.

### Visualizing the Power Problem

```{r power-plot, fig.height=5.5}
plot_round1 <- shoe_data |>
  filter(cookie == "F") |>
  mutate(size = factor(size,
                       levels = c("Small", "Large"),
                       labels = c("n = 5 each", "n = 100 each")))

ann_round1 <- tibble(
  size  = factor(c("n = 5 each", "n = 100 each"),
                 levels = c("n = 5 each", "n = 100 each")),
  label = c(
    as.character(sprintf("p = %.3f (ns)", p_small)),
    as.character(sprintf("p = %s ***",
                         format.pval(p_large, digits = 2, eps = 0.001)))
  ),
  y_pos = c(max(plot_round1$ci_hi[plot_round1$size == "n = 5 each"]) + 12,
            max(plot_round1$ci_hi[plot_round1$size == "n = 100 each"]) + 8)
)

ggplot(plot_round1, aes(x = shoe, y = mean, fill = shoe)) +
  geom_col(width = 0.6, alpha = 0.85) +
  geom_errorbar(aes(ymin = ci_lo, ymax = ci_hi),
                width = 0.2, linewidth = 0.7) +
  geom_signif(
    data = ann_round1 |> filter(size == "n = 5 each"),
    aes(xmin = "Men's", xmax = "Women's",
        annotations = label, y_position = y_pos),
    inherit.aes = FALSE, manual = TRUE, tip_length = 0.02
  ) +
  geom_signif(
    data = ann_round1 |> filter(size == "n = 100 each"),
    aes(xmin = "Men's", xmax = "Women's",
        annotations = label, y_position = y_pos),
    inherit.aes = FALSE, manual = TRUE, tip_length = 0.02
  ) +
  facet_wrap(~ size) +
  scale_fill_manual(values = shoe_palette) +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(
    x = NULL, y = "Mean price",
    title = "Same store, same cookie, same gender gap",
    subtitle = "Five shoes can't see what 100 shoes can. Bars show 95% CIs."
  ) +
  theme(legend.position = "none")
```

Bars are similar heights in both panels. What changes is the width of the
confidence intervals. Five shoes give you huge intervals that overlap massively.
A hundred shoes give you tight intervals that clearly separate. The data didn't
change. Our ability to see the data did.

## Round 2: One-Sample *z*-Tests

We have the population standard deviations, so a *z*-test is appropriate. Three
questions about the curated samples we got served.

```{r z-helper}
one_sample_z <- function(x, mu, sigma) {
  n  <- length(x)
  m  <- mean(x)
  se <- sigma / sqrt(n)
  z  <- (m - mu) / se
  p  <- 2 * (1 - pnorm(abs(z)))
  tibble(M = m, n = n, mu = mu, sigma = sigma,
         z = z, p = p, se = se)
}
```

### Test 3: Is the Curated Men's Sample Weird Compared to All Men's Shoes?

```{r z1}
z_men_vs_men <- one_sample_z(mens_small_F, mu_men, sigma_men)
z_men_vs_men
```

*z* = `r round(z_men_vs_men$z, 2)`,
*p* = `r format.pval(z_men_vs_men$p, digits = 2, eps = 0.001)`. The five "men's
shoes" the algorithm picked for the female-cookie shopper aren't a normal slice
of the men's inventory. They're significantly more expensive than the average
men's shoe.

### Test 4: Does That Same Curated Sample Look Like Women's Shoes?

```{r z2}
z_men_vs_women <- one_sample_z(mens_small_F, mu_women, sigma_women)
z_men_vs_women
```

*z* = `r round(z_men_vs_women$z, 2)`,
*p* = `r round(z_men_vs_women$p, 3)`. The "men's" shoes shown to the
female-cookie shopper are priced indistinguishably from the women's
population. The algorithm wasn't picking random men's shoes. It was picking
men's shoes that cost what women's shoes cost.

### Test 5: Sanity Check on the Large Women's Sample

```{r z3}
z_women_check <- one_sample_z(womens_large_F, mu_women, sigma_women)
z_women_check
```

*z* = `r round(z_women_check$z, 2)`,
*p* = `r round(z_women_check$p, 3)`. The large women's sample sits comfortably
near the women's population mean. When we sample broadly, we recover the
population. The bias lives in the small, curated picks.

### Visualizing Where the Curated Sample Falls

```{r z-plot, fig.height=5}
x_grid <- seq(40, 150, length.out = 400)
sample_mean_curves <- bind_rows(
  tibble(price = x_grid,
         density = dnorm(x_grid, mu_men, sigma_men / sqrt(5)),
         hypothesis = "If men's pop"),
  tibble(price = x_grid,
         density = dnorm(x_grid, mu_women, sigma_women / sqrt(5)),
         hypothesis = "If women's pop")
)

curated_mean <- mean(mens_small_F)

ggplot(sample_mean_curves, aes(x = price, y = density, color = hypothesis)) +
  geom_line(linewidth = 1.1) +
  geom_vline(xintercept = curated_mean, linetype = "dashed",
             color = "#222222", linewidth = 0.7) +
  annotate("label", x = curated_mean, y = 0.058,
           label = sprintf("Curated men's\nsample mean = $%.2f", curated_mean),
           hjust = 0.5, size = 3.3, fill = "white") +
  annotate("text", x = mu_men - 8, y = 0.045,
           label = sprintf("z = %.2f\np = %s",
                           z_men_vs_men$z,
                           format.pval(z_men_vs_men$p, digits = 2, eps = 0.001)),
           color = shoe_palette[["Men's"]], size = 3.3, fontface = "bold") +
  annotate("text", x = mu_women + 12, y = 0.025,
           label = sprintf("z = %.2f\np = %.3f",
                           z_men_vs_women$z, z_men_vs_women$p),
           color = shoe_palette[["Women's"]], size = 3.3, fontface = "bold") +
  scale_color_manual(values = c("If men's pop"   = shoe_palette[["Men's"]],
                                "If women's pop" = shoe_palette[["Women's"]]),
                     name = "Sampling distribution") +
  scale_x_continuous(labels = scales::dollar_format()) +
  labs(
    x = "Sample mean ($, n = 5)", y = "Density",
    title = "The curated sample sits far from men's, comfortable with women's",
    subtitle = "Sampling distributions of the mean for n = 5"
  ) +
  theme(legend.position = "bottom")
```

The dashed line marks the curated sample mean of
\$`r round(curated_mean, 2)`. Look how far it sits in the right tail of the
men's sampling distribution, and how comfortably it sits inside the women's
sampling distribution. The *z*-tests are reading off where that line falls.

## Round 3: The Plot Twist (Cookie = M)

We cleared cookies, browsed a few hardware sites, and came back. Now the
algorithm thinks we're male. We refreshed the same shoes.

```{r cookie-comparison}
cookie_table <- shoe_data |>
  select(shoe, cookie, size, mean) |>
  pivot_wider(names_from = cookie, values_from = mean,
              names_prefix = "Cookie = ") |>
  mutate(`Difference (F - M)` = `Cookie = F` - `Cookie = M`)

cookie_table
```

Female-cookie prices come in higher, and the gap is huge for the small curated
samples. The right test for this is paired, since each cookie=F price has a
matched cookie=M price for the same physical shoe.

## Round 4: Paired *t*-Test

For the paired test, we use the differences directly. To make this work with
our simulation, we generate per-shoe price differences for the same 100 men's
shoes and the same 100 women's shoes across cookie conditions.

```{r paired-t}
set.seed(43)

mens_diff   <- rnorm(100, mean = 3.41, sd = 9.20)
womens_diff <- rnorm(100, mean = 0.41, sd = 4.51)

paired_t <- function(diffs) {
  n        <- length(diffs)
  mean_d   <- mean(diffs)
  sd_d     <- sd(diffs)
  se       <- sd_d / sqrt(n)
  t_stat   <- mean_d / se
  df       <- n - 1
  p_val    <- 2 * (1 - pt(abs(t_stat), df))
  tibble(mean_diff = mean_d, sd_diff = sd_d, n = n,
         t = t_stat, df = df, p = p_val, se = se)
}

paired_men   <- paired_t(mens_diff)
paired_women <- paired_t(womens_diff)

bind_rows(
  paired_men   |> mutate(comparison = "Men's shoes (F vs M cookie)"),
  paired_women |> mutate(comparison = "Women's shoes (F vs M cookie)")
) |>
  select(comparison, everything())
```

For the men's shoes, the average price drop after switching cookies was
\$`r round(paired_men$mean_diff, 2)`,
*t*(`r paired_men$df`) = `r round(paired_men$t, 2)`,
*p* = `r format.pval(paired_men$p, digits = 2, eps = 0.001)`. For the women's
shoes, the average drop was \$`r round(paired_women$mean_diff, 2)`,
*t*(`r paired_women$df`) = `r round(paired_women$t, 2)`,
*p* = `r round(paired_women$p, 3)`. The cookie effect is real for men's shoes
and absent for women's shoes. The store appears to be charging female-coded
shoppers more specifically when they shop in the men's section.

```{r paired-plot, fig.height=4.8}
paired_plot_data <- shoe_data |>
  filter(size == "Large") |>
  select(shoe, cookie, mean, ci_lo, ci_hi)

paired_label_men <- sprintf(
  "Men's: t(%d) = %.2f, p = %s",
  paired_men$df, paired_men$t,
  format.pval(paired_men$p, digits = 2, eps = 0.001)
)
paired_label_women <- sprintf(
  "Women's: t(%d) = %.2f, p = %.3f",
  paired_women$df, paired_women$t, paired_women$p
)

ggplot(paired_plot_data,
       aes(x = cookie, y = mean, group = shoe, color = shoe)) +
  geom_line(linewidth = 1.2) +
  geom_errorbar(aes(ymin = ci_lo, ymax = ci_hi),
                width = 0.08, linewidth = 0.7) +
  geom_point(size = 4) +
  geom_text(aes(label = scales::dollar(mean, accuracy = 0.01)),
            nudge_y = 1.5, size = 3.5, show.legend = FALSE) +
  annotate("text", x = 1.5, y = min(paired_plot_data$ci_lo) - 2,
           label = paired_label_men,
           size = 3.5, color = shoe_palette[["Men's"]], fontface = "bold") +
  annotate("text", x = 1.5, y = max(paired_plot_data$ci_hi) + 4,
           label = paired_label_women,
           size = 3.5, color = shoe_palette[["Women's"]], fontface = "bold") +
  scale_color_manual(values = shoe_palette) +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(
    x = "Tracking cookie", y = "Mean price",
    title = "The cookie effect is selective",
    subtitle = "Men's prices drop when the cookie says male. Women's prices barely move."
  ) +
  theme(legend.position = "bottom")
```

The men's line has a visible slope. The women's line is nearly flat. The
tracking cookie is changing prices in a targeted way.

## Round 5: One-Way ANOVA

Four groups: each combination of shoe gender and cookie status. A one-way
ANOVA treats them as four conditions and asks whether at least one mean differs
from the others.

```{r anova}
anova_data <- bind_rows(
  tibble(price = mens_large_F,   group = "Men's, Cookie=F"),
  tibble(price = mens_large_M,   group = "Men's, Cookie=M"),
  tibble(price = womens_large_F, group = "Women's, Cookie=F"),
  tibble(price = womens_large_M, group = "Women's, Cookie=M")
) |>
  mutate(group = factor(group,
                        levels = c("Men's, Cookie=F", "Men's, Cookie=M",
                                   "Women's, Cookie=F", "Women's, Cookie=M")))

anova_fit     <- aov(price ~ group, data = anova_data)
anova_summary <- summary(anova_fit)[[1]]
anova_F       <- anova_summary[["F value"]][1]
anova_p       <- anova_summary[["Pr(>F)"]][1]
df_btw        <- anova_summary[["Df"]][1]
df_wn         <- anova_summary[["Df"]][2]

summary(anova_fit)
```

*F*(`r df_btw`, `r df_wn`) = `r round(anova_F, 2)`,
*p* = `r format.pval(anova_p, digits = 2, eps = 0.001)`. At least one of the
four conditions has a different mean than the others. The ANOVA on its own
won't tell us which pairs differ. We can see the pattern in the means though.

```{r anova-plot, fig.height=5.5}
anova_plot_data <- shoe_data |>
  filter(size == "Large") |>
  mutate(group = paste0(shoe, ",\nCookie=", cookie),
         group = factor(group,
                        levels = c("Men's,\nCookie=F", "Men's,\nCookie=M",
                                   "Women's,\nCookie=F", "Women's,\nCookie=M")))

anova_label <- sprintf(
  "Overall ANOVA: F(%d, %d) = %.2f, p = %s",
  df_btw, df_wn, anova_F,
  format.pval(anova_p, digits = 2, eps = 0.001)
)

paired_men_label <- sprintf(
  "Men's: F vs M cookie, p = %s",
  format.pval(paired_men$p, digits = 2, eps = 0.001)
)
paired_women_label <- sprintf(
  "Women's: F vs M cookie, p = %.3f",
  paired_women$p
)

y_top <- max(anova_plot_data$ci_hi)

ggplot(anova_plot_data, aes(x = group, y = mean, fill = shoe)) +
  geom_col(width = 0.65, alpha = 0.85) +
  geom_errorbar(aes(ymin = ci_lo, ymax = ci_hi),
                width = 0.18, linewidth = 0.7) +
  geom_text(aes(label = scales::dollar(mean, accuracy = 0.01)),
            vjust = -0.6, size = 3.4, fontface = "bold") +
  geom_signif(annotations = paired_men_label,
              xmin = "Men's,\nCookie=F", xmax = "Men's,\nCookie=M",
              y_position = y_top + 6, tip_length = 0.02, textsize = 3.2) +
  geom_signif(annotations = paired_women_label,
              xmin = "Women's,\nCookie=F", xmax = "Women's,\nCookie=M",
              y_position = y_top + 14, tip_length = 0.02, textsize = 3.2) +
  annotate("label", x = 2.5, y = y_top + 26,
           label = anova_label,
           fill = "#FFF3B0", size = 3.6, fontface = "bold") +
  scale_fill_manual(values = shoe_palette) +
  scale_y_continuous(labels = scales::dollar_format(),
                     limits = c(0, y_top + 35)) +
  labs(
    x = NULL, y = "Mean price",
    title = "Four groups, one ANOVA, a clear pattern",
    subtitle = "95% CIs shown. Brackets are paired comparisons from Round 4."
  ) +
  theme(legend.position = "none",
        axis.text.x = element_text(size = 9))
```

Two patterns show up in the bars. Women's shoes are taller than men's shoes
regardless of cookie. Within men's shoes, the female-cookie bar is taller than
the male-cookie bar. Within women's shoes, the cookie barely matters.

## So, Is the Store Gouging?

Pulling everything together:

- The **large-sample independent *t*-test** showed women's shoes really do cost
  more than men's shoes,
  *t*(`r df_large`) = `r round(t_stat_large, 2)`,
  *p* = `r format.pval(p_large, digits = 2, eps = 0.001)`.
- The ***z*-tests** showed the curated small sample of men's shoes shown to a
  female-cookie shopper was priced up in women's-shoe territory
  (*z* = `r round(z_men_vs_men$z, 2)` against the men's population,
  *z* = `r round(z_men_vs_women$z, 2)` against the women's population).
- The **paired *t*-test** showed that on the same men's shoes, female-cookie
  shoppers pay about \$`r round(paired_men$mean_diff, 2)` more than male-cookie
  shoppers. For women's shoes, the cookie does nothing
  (*p* = `r round(paired_women$p, 3)`).
- The **one-way ANOVA** confirmed the four shoe-by-cookie conditions aren't
  drawn from the same distribution,
  *F*(`r df_btw`, `r df_wn`) = `r round(anova_F, 2)`,
  *p* = `r format.pval(anova_p, digits = 2, eps = 0.001)`.

The hypothesis isn't proven beyond doubt. A real audit would need a controlled
experiment with many more shoes, randomized browsing, and a few different IP
addresses. The evidence we have is consistent and points one direction though.
The cookie matters, and sample size matters even more. With five shoes, none of
this would have been visible.

::: {.callout-tip}
## The Big Lesson
A small biased sample can perfectly hide a large biased system. Always ask how
much data a conclusion came from.
:::

::: {.callout-note}
## For the Review Session
Bring questions about confidence intervals. Every plot in this document shows
them, and we'll spend time talking about what a 95% CI does and does not say.
:::
